# dplyr pacakage in R

Siyu Shen  


The **dplyr** package is mainly used for cleaning and sorting data. By using **dplyr** package, we can efficiently process the data, which is very important for people to do data analysis. In this note, I am going to introduce the most common used functions in **dplyr**.

```{r eval=FALSE, message=FALSE, warning=FALSE}
#install 'dplyr' package
install.packages(dplyr)
```

```{r echo=T, message=FALSE, results='hide'}
#load 'dplyr' package
library(dplyr)
```

```{r}
#load data 'mtcars' as example
data(mtcars)

#show the structure of the data 'mtcars'
str(mtcars)
```

There are 32 observations on 11 variables. The meanings of each variables are:  
**mpg**: Miles/(US) gallon,  
**cyl**: Number of cylinders,  
**disp**:	Displacement (cu.in.),  
**hp**:	Gross horsepower,  
**drat**:	Rear axle ratio,  
**wt**:	Weight (1000 lbs),  
**qsec**:	1/4 mile time,  
**vs**:	Engine (0 = V-shaped, 1 = straight),  
**am**:	Transmission (0 = automatic, 1 = manual),  
**gear**:	Number of forward gears,  
**carb**:	Number of carburetors.  

## 1. Pipe operator '%>%'

It is important to know the pipe **%>%** operator before we start to learn the functions in **dplyr** package. With the use of **%>%** , multiple functions can be wrapper together. And it can be used with any function.

```{r eval=FALSE}
# usage in filter() function
filter(data_frame, variable == value)
data_frame %>% filter(variable == value)

# usage in mutate() function
mutate(data_frame, expression(s))
data_frame %>% mutate(expression(s))
```


## 2. Filter by row: filter() function

You can use **filter()** function to filter the subsets by the given logic, which is similar to the **subset()** function, for example:

```{r}
#filter when mpg is greater than or equal to 22
filter(mtcars, mpg >= 22)
```

```{r eval=FALSE}
#filter when cyl=4 or gear=3
filter(mtcars, cyl == 4 | gear == 3)

#filter when cyl=4 and gear=3
filter(mtcars, cyl == 4 & gear == 3)

#note: when using AND operation, avoid using '&&' instead of '&'
```


## 3. Filter by column: select() function

**select()** function selects subdatasets with column names as arguments. The **dplyr** package provides some special functions to be used in conjunction with the **select()** function to filter variables, including **starts_with**, **ends_with**, **contains**, **matches**, **one_of**, **num_range**, and **everything**.

```{r warning=FALSE}
#choose data 'iris' as example
data(iris)

iris = tbl_df(iris)

#show the 'iris' data
head(iris)
```

```{r}
#select the columns that start with 'Petal'
select(iris, starts_with("Petal"))
```

```{r eval=FALSE}
#select the columns that not start with 'Petal'
select(iris, -starts_with("Petal"))
select(iris, !starts_with("Petal"))

#select the columns that end with 'Petal'
select(iris, ends_with("Width"))

#select the columns that contain with 'etal'
select(iris, contains("etal"))

#select the columns that the variables name is with 't'
select(iris, matches(".t."))

#select the columns you want directly
select(iris, Petal.Length, Petal.Width)

#select multi-columns by using colon
select(iris, Sepal.Length:Petal.Width)
```

```{r eval=FALSE}
#when we cannot use the character vector filter, we need to use 'one_of()' function
vars <- c("Petal.Length", "Petal.Width")
select(iris, one_of(vars))
```

```{r eval=FALSE}
#return all columns, generally used when adjusting the order of variables in a dataset
select(iris, everything())

#return all columns, but reorder the columns and put 'Species' column in the front
select(iris, Species, everything())
```

## 4. mutate() and transmute()

**mutate()** adds new variables and preserves existing ones; **transmute()** adds new variables and drops existing ones.  

**4.1. mutate()**  

**4.1.1. New columns**  

```{r}
#create two more columns named 'cyl2' and 'cyl4'
head(mtcars %>% mutate(cyl2 = cyl * 2, cyl4 = cyl2 * 2))
```

**4.1.2. Delete columns**  

```{r eval=FALSE}
#delete the column 'mpg' and update the column 'disp'
mtcars %>% mutate(mpg = NULL, disp = disp * 0.0163871)

#delete the column 'cyl'
mtcars %>% mutate(cyl = NULL)
```

**4.1.3. Application of Window function**  

```{r}
#create a new column 'rank' using min_rank() function by group 'cyl'
head(mtcars %>% group_by(cyl) %>% mutate(rank = min_rank(desc(mpg))))

#create a new column 'mpg_max' using max() function by group 'cyl'
head(mtcars %>% group_by(cyl) %>% mutate(mpg_max = max(mpg)))

```
  

**4.2. transmute()**  

The return value does not contain the original dataset variables, only the variables after calculation and transformation are retained.  

```{r}
head(mtcars %>% mutate(wt_log=log(wt)))

head(mtcars %>% transmute(wt_log=log(wt)))

head(mtcars %>% mutate(displ_l = disp / 61.0237))

head(mtcars %>% transmute(displ_l = disp / 61.0237))
```

## 5. Ranking function

row_number: the results of the parallel rankings are in different order, and the elements appearing first are ranked first.  

min_rank: the results of the tie ranks are the same, and the next rank will be occupied.  

dense_rank: tied ranking does not occupy the ranking, for example: no matter how many tied for the second place, the subsequent ranking should still be the third place  

percent_rank: rank by percentage  

cume_dist: rank of cumulative distribution interval  

ntile: roughly ranks vectors by dividing into n buckets. Larger buckets have lower rank.

```{r}
x = c(5, 1, 3, 2, 2, NA)

row_number(x)

min_rank(x)

dense_rank(x)

percent_rank(x)

cume_dist(x)

ntile(x, 2)

head(mtcars%>%mutate(dense_rank=cume_dist(cyl)))
```

## 6. Sort function: arrange()  

Note the difference between sorting and ranking.  

**arrange()** sorts the rows sequentially by the given column names.  

```{r}
#sorted by column 'mpg'
head(arrange(mtcars, mpg))

#sorted by column 'mpg' and 'disp'
head(arrange(mtcars, mpg, disp))

#reverse order by using desc()
head(arrange(mtcars, desc(mpg)))

#reverse order by using '-'
head(arrange(mtcars, -mpg))
```

## 7. Summary function: summarise()

**7.1. Summary in whole**

```{r}
#return the mean value in column 'disp'
summarise(mtcars, mean(disp))

#return the standard deviation in column 'disp'
summarise(mtcars, sd(disp))

#return the maximum and minimum value in column 'disp'
summarise(mtcars, max(disp), min(disp))

#return the number of columns in dataset 'mtcars'
summarise(mtcars, n())

#return the number of unique gear
summarise(mtcars, n_distinct(gear))

#return the first value in column 'disp'
summarise(mtcars, first(disp))

#return the last value in column 'disp'
summarise(mtcars, last(disp))
```

**7.2. Summary by group**

```{r}
#return the mean value and standard deviation of column 'disp' by grouping column 'cyl'
mtcars %>% group_by(cyl) %>%
summarise(mean = mean(disp), sd = sd(disp))
```

