# Interactive 3D Visualization in R

Yaoyuan Zhang and Zimu An

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(webgl = "hook_webgl")
#sdevtools::install_github("hypertidy/anglr") #Run this line if not installed before
library(plotly)
library(rgl)
library(anglr)
library(maptools)
```

## Motivation

Although a 2D plot (or multiple of them) can contain enough information
we need, a 3D plot can be more intuitive since the 3D space is where we
reside. For example, we usually see the 2D projection of a map, or
atlas, in textbooks or websites, but a globe is always more pleasant to
use. In addition, when encoutering datasets with higher dimensions in
statistics, a 3D visualization can help us with generalizing in higher
dimensional spaces.

In this tutorial, we focus on interactive 3D visualization using packages including 
`plotly` and `rgl`. We first cover the basic 3D scatter plots using the package `rgl`, 
then we try to add more useful tools for better visualization. We also introduce
how to produce interactive 3D surface graphs and 3D mesh plots in R
using the package `plotly`

We hope this tutorial can provide some insights and inspirations for
people who wish to tackle high-dimensional data visualization problems.

## 3D visualization with `rgl` package

The `rgl` package is used to produce interactive 3-D plots. It contains
high-level graphics commands modeled loosely after classic R graphics,
but working in three dimensions. It also contains low level structure
inspired by (but incompatible with) the `grid` package.

### Preparing the Data

We'll use the `iris` data for this tutorial in the following examples.

`iris` data set gives the measurements of the variables sepal length and
width, petal length and width, respectively, for 50 flowers from each of
3 species of iris. The species are Iris setosa, versicolor, and
virginica.

```{r}
data(iris)
head(iris)
```

```{r}
x <- sep.l <- iris$Sepal.Length
y <- pet.l <- iris$Petal.Length
z <- sep.w <- iris$Sepal.Width
```

### Starting and Closing the RGL Device

In order to make a 3D plot with RGL, we need to first start the RGL
device in R. The functions below are used to manage RGL device:

-   **rgl.open()**: Opens a new device
-   **rgl.close()**: Closes the current device
-   **rgl.clear()**: Clears the current device
-   **rgl.cur()**: Returns the active device ID
-   **rgl.quit()**: Shutdowns the RGL device system

### 3D Scatter Plot

#### Custom function to initialize RGL device

Since the default RGL device usually needs initialization, we can create
a custom function for easier initialization process. The following
function **rgl_init()** will create a new **RGL** device if there is no
opened device:

```{r}
#' @param new.device a logical value. If TRUE, creates a new device
#' @param bg the background color of the device
#' @param width the width of the device
rgl_init <- function(new.device = FALSE, bg = "white", width = 640) { 
  if( new.device | rgl.cur() == 0 ) {
    rgl.open()
    par3d(windowRect = 50 + c( 0, 0, width, width ) )
    rgl.bg(color = bg )
  }
  rgl.clear(type = c("shapes", "bboxdeco"))
  rgl.viewpoint(theta = 15, phi = 20, zoom = 0.7)
}
```

The functions used in the custom function that were not introduced
previously includes:

-   **par3d(windowRect)**: set the window size
-   **rgl.viewpoint(theta, phi, fov, zoom)**: set up viewpoint. The
    arguments *theta* and *phi* are polar coordinates.
    -   **theta** and **phi** are polar coordinates. Default values are
        0 and 15, respectively
    -   **fov** is the field-of-view angle in degrees. Default value is
        60
    -   **zoom** is the zoom factor. Default value is 1
-   **rgl.bg(color)**: define the background color of the device

#### Basic 3D Plot with spheres

```{r, test-rgl, webgl=TRUE}
rgl_init()
rgl.spheres(x, y, z, r = 0.1, color = "yellow")  # Scatter plot
```

Note that `rgl` package also provides functions such as **rgl.phoints**
to represent data points in another format. For the purpose of
conformity, this tutorial use spheres to represent data points.

#### Add a Bounding Box decoration

The function **rgl.bbox()** is used:

```{r, test-rgl, webgl=TRUE}
rgl_init()
rgl.spheres(x, y, z, r = 0.1, color = "yellow")  
# Add bounding box decoration
rgl.bbox(color=c("#333377","black"), emission="#333377",
         specular="#3333FF", shininess=5, alpha=0.8 ) 
```

-   **color**: a vector of colors. The first color is used for the
    background color of the bounding box. The second color is used for
    the tick mark labels.
-   **emission**, **specular**, **shininess**: properties for lighting
    calculation
-   **alpha**: value specifying the color transparency. The value should
    be between 0.0 (fully transparent) and 1.0 (opaque)

#### Add axis lines and labels

-   The function **rgl.lines(x, y = NULL, z = NULL, ...)** can be used
    to add axis lines.
-   The function **rgl.texts(x, y = NULL, z = NULL, text)** is used to
    add axis labels.

For the convenience of plotting, a custom function **rgl_add_axes()** is
implemented for the purpose of adding x, y, and z axes.

```{r}
# x, y, z : numeric vectors corresponding to
#  the coordinates of points
# axis.col : axis colors
# xlab, ylab, zlab: axis labels
# show.plane : add axis planes
# show.bbox : add the bounding box decoration
# bbox.col: the bounding box colors. The first color is the
# the background color; the second color is the color of tick marks
rgl_add_axes <- function(x, y, z, axis.col = "grey",
                xlab = "", ylab="", zlab="", show.plane = TRUE, 
                show.bbox = FALSE, bbox.col = c("#333377","black"))
  { 
  
  lim <- function(x){c(-max(abs(x)), max(abs(x))) * 1.1}
  # Add axes
  xlim <- lim(x); ylim <- lim(y); zlim <- lim(z)
  rgl.lines(xlim, c(0, 0), c(0, 0), color = axis.col)
  rgl.lines(c(0, 0), ylim, c(0, 0), color = axis.col)
  rgl.lines(c(0, 0), c(0, 0), zlim, color = axis.col)
  
   # Add a point at the end of each axes to specify the direction
   axes <- rbind(c(xlim[2], 0, 0), c(0, ylim[2], 0), 
                 c(0, 0, zlim[2]))
   rgl.points(axes, color = axis.col, size = 3)
  
  # Add axis labels
  rgl.texts(axes, text = c(xlab, ylab, zlab), color = axis.col,
             adj = c(0.5, -0.8), size = 2)
  
  # Add plane
  if(show.plane) 
    xlim <- xlim/1.1; zlim <- zlim /1.1
    rgl.quads( x = rep(xlim, each = 2), y = c(0, 0, 0, 0),
             z = c(zlim[1], zlim[2], zlim[2], zlim[1]))
  
  # Add bounding box decoration
  if(show.bbox){
    rgl.bbox(color=c(bbox.col[1],bbox.col[2]), alpha = 0.5, 
          emission=bbox.col[1], specular=bbox.col[1], shininess=5, 
          xlen = 3, ylen = 3, zlen = 3) 
  }
}
```

-   **rgl.quads(x, y, z)** is used to add planes. x, y and z are numeric
    vectors of length four specifying the coordinates of the four nodes
    of the quad.

```{r, test-rgl, webgl=TRUE}
rgl_init()
rgl.spheres(x, y, z, r = 0.1, color = "yellow") 
rgl_add_axes(x, y, z, show.bbox = TRUE)
# Show tick marks
axis3d('x', pos=c( NA, 0, 0 ), col = "darkgrey")
axis3d('y', pos=c( 0, NA, 0 ), col = "darkgrey")
axis3d('z', pos=c( 0, 0, NA ), col = "darkgrey")
```

The function **axis3d()** is used in the previous block to show the
scales of axes.

#### Set aspect ratios of the axes

The function **aspect3d(x, y = NULL, z = NULL)** can be used to set the
apparent ratios of the x, y and z axes for the current plot.

x, y and z are the ratio for x, y and z axes, respectively. x can be a
vector of length 3, specifying the ratio for the 3 axes.

#### Change color of points by groups

A helper function can be used to select automatically a color for each
group:

```{r}
#' Get colors for the different levels of 
#' a factor variable
#' 
#' @param groups a factor variable containing the groups
#'  of observations
#' @param colors a vector containing the names of 
#   the default colors to be used
get_colors <- function(groups, group.col = palette()){
  groups <- as.factor(groups)
  ngrps <- length(levels(groups))
  if(ngrps > length(group.col)) 
    group.col <- rep(group.col, ngrps)
  color <- group.col[as.numeric(groups)]
  names(color) <- as.vector(groups)
  return(color)
}
```

Change colors by groups:

```{r, test-rgl, webgl=TRUE}
cols <- get_colors(iris$Species, c("lightblue", "palegreen3", "slateblue3"))
rgl_init()
rgl.spheres(x, y, z, r = 0.1, color = cols) 
rgl_add_axes(x, y, z, show.bbox = TRUE)
aspect3d(1,1,1)
```

### Additional Analysis tools

#### Ellipse of Concentration

The function **ellipse3d()** is used to estimate the ellipse of
concentration:

-   **x**: the correlation or covariance matrix between x, y and z
-   **scale**: If x is a correlation matrix, then the standard
    deviations of each parameter can be given in the scale parameter.
    This defaults to c(1, 1, 1), so no rescaling will be done.
-   **centre**: The center of the ellipse will be at this position.
-   **level**: The confidence level of a confidence region. This is used
    to control the size of the ellipsoid.

The function **ellipse3d()** returns an object of class *mesh3d* which
can be drawn using the function **shade3d()** and/or **wired3d()**

Draw the ellipse using the function **shade3d()**:

```{r, test-rgl, webgl=TRUE}
rgl_init()
rgl.spheres(x, y, z, r = 0.12, color = "#D95F02") 
rgl_add_axes(x, y, z, show.bbox = TRUE)
# Compute and draw the ellipse of concentration
ellips <- ellipse3d(cov(cbind(x,y,z)), 
            centre=c(mean(x), mean(y), mean(z)), level = 0.95)
shade3d(ellips, col = "#D95F02", alpha = 0.1, lit = FALSE)
aspect3d(1,1,1)
```

Draw the ellipse using the function **wired3d()**:

```{r, test-rgl, webgl=TRUE}
rgl_init()
rgl.spheres(x, y, z, r = 0.2, color = "#D95F02") 
rgl_add_axes(x, y, z, show.bbox = TRUE)
# Compute and draw the ellipse of concentration
ellips <- ellipse3d(cov(cbind(x,y,z)), 
            centre=c(mean(x), mean(y), mean(z)), level = 0.95)
wire3d(ellips, col = "#D95F02",  lit = FALSE)
aspect3d(1,1,1)
```

We can also add ellipse for each group, as such:

```{r, test-rgl, webgl=TRUE}
# Groups
groups <- iris$Species
levs <- levels(groups)
group.col <- c("red", "green", "blue")
# Plot observations
rgl_init()
rgl.spheres(x, y, z, r = 0.1,
            color = group.col[as.numeric(groups)]) 
rgl_add_axes(x, y, z, show.bbox = FALSE)
# Compute ellipse for each group
for (i in 1:length(levs)) {
    group <- levs[i]
    selected <- groups == group
    xx <- x[selected]; yy <- y[selected]; zz <- z[selected]
    ellips <- ellipse3d(cov(cbind(xx,yy,zz)), 
              centre=c(mean(xx), mean(yy), mean(zz)), level = 0.95) 
    shade3d(ellips, col = group.col[i], alpha = 0.1, lit = FALSE) 
    # show group labels
    texts3d(mean(xx),mean(yy), mean(zz), text = group,
            col= group.col[i], cex = 1)
  }
aspect3d(1,1,1)
```

#### Regression plane

To add a custom regression plane, we can take the following approach:

1.  Use the function **lm()** to compute a linear regression model.
2.  Use the argument **rgl.surface()** to add a regression surface.

```{r, test-rgl, webgl=TRUE}
rgl_init()
rgl.spheres(x, y, z, r = 0.1, color = "#D95F02") 
rgl_add_axes(x, y, z, show.bbox = FALSE)
aspect3d(1,1,1)
# Compute the linear regression (y = ax + bz + d)
fit <- lm(y ~ x + z)
# predict values on regular xz grid
grid.lines = 26
x.pred <- seq(min(x), max(x), length.out = grid.lines)
z.pred <- seq(min(z), max(z), length.out = grid.lines)
xz <- expand.grid( x = x.pred, z = z.pred)
y.pred <- matrix(predict(fit, newdata = xz), 
                 nrow = grid.lines, ncol = grid.lines)
# Add regression surface
rgl.surface(x.pred, z.pred, y.pred, color = "steelblue", 
                alpha = 0.5, lit = FALSE)  
# Add grid lines
rgl.surface(x.pred, z.pred, y.pred, color = "black",
    alpha = 0.5, lit = FALSE, front = "lines", back = "lines")
```

## Interactive 3D surface plots with package `plotly`

### 3D Surface Plot

In this section, we will introduce how to use plotly to generate interactive 3D
surface plots. 3D suface plots are extremely useful and can convey/visualize a
lot more information than a traditional contour plot, especially towards topological,
geographical data in a straight forward fashion.

The dataset we are using here is the 'volcano' (Topographic Information on
Auckland's Maunga Whau Volcano) dataset. Maunga Whau (Mt Eden) is one of about
50 volcanos in the Auckland volcanic field. This data set gives topographic
information for Maunga Whau on a 10m by 10m grid. The dataset is a matrix with
87 rows and 61 columns, rows corresponding to grid lines running east to west
and columns to grid lines running south to north. More importantly, since it's
a numeric matrix we could easily take it as an argument for method **plot_ly**
which initializes the instance.
```{r}
head(volcano)
```


```{r}
fig <- plot_ly(z = ~volcano)
fig <- fig %>% add_surface()

fig
```
Just few simple lines, we can create a highly interactive 3D plot that
illustrates the 3D representation of the volcano dataset. The lighter the color,
the higher the altitude. By rolling the mouse, we can zoom in and out easily.
Pointing the mouse at every vertex on this 3D graph, the user can easily see the
x, y, z coordinates while also the sectional view. From the above, we can also
see the contour lines clearly displayed.

Moreover, we can configure the parameters in **add_surface()**. Here, we will
modify the contours argument to display contour lines at the top/below the 3D
graph depending on the view of the user. By setting highlight color to be red and
setting project, when we move our cursor to each contour line, it maps the
contour directly to the graph.

We can approach almost any other numeric matrices that displays geographical
information using this package. With the smooth transformation, animation, and
easy-to-use convenience, it would largely improve users' experience.

```{r}
fig <- plot_ly(z = ~volcano) %>% add_surface(
  # configuring contours mapping
  contours = list(
    z = list(
      show=TRUE,
      usecolormap=TRUE,
      highlightcolor="red",
      project=list(z=TRUE)
      )
    )
  )

fig
```
Other than geographical data, we can also apply the approach on kernel density.
Here we are using the 'Old Faithful Geyser Data', which is a version of the
eruptions data from the ‘Old Faithful’ geyser in Yellowstone National Park,
Wyoming. The dataset has 299 observations on 2 variables: duration (numeric
eruption time in minutes) and waiting (numeric waiting time for this eruption).

We used kde2d to take duration as the x-coordinate, waiting as y-coordinate, and
produce a Two-dimensional kernel density estimation, where z is the estimation.


```{r}
kd <- with(MASS::geyser, MASS::kde2d(duration, waiting, n = 50))
# plot using plotly by taking the kernel density created
fig <- plot_ly(x = kd$x, y = kd$y, z = kd$z) %>% add_surface(
  contours = list(
    z = list(
      show=TRUE,
      usecolormap=TRUE,
      highlightcolor="red",
      project=list(z=TRUE)
      )
    )
  ) %>% layout(
    # label the axis using layout
    scene = list(
                xaxis = list(title = "duration"),
                yaxis = list(title = "waiting" ),  
                zaxis = list(title = "Estimate")
           )
  )

fig
```

### Tri-surface Plot

A **Tri-Surface Plot** is a type of surface plot, created by
triangulation of compact surfaces of finite number of triangles which
cover the whole surface in a manner that each and every point on the
surface is in triangle. The intersection of any two triangles results in
void or a common edge or vertex. This type of plot is created where the
evenly sampled grids are restrictive and inconvenient to plot.

#### Basic Tri-Surf plot using `plotly`

Since the tri-surface plot is constructed with triangular surfaces, we
start with the simplest geometry object to plot in 3D - a regular
tetrahedron:

```{r, warning=FALSE}
suppressPlotlyMessage <- function(p) {
  suppressMessages(plotly_build(p))
}

suppressPlotlyMessage(plot_ly(
  x = c(0, 1, 2, 0),
  y = c(0, 0, 1, 2),
  z = c(0, 2, 0, 1),
  i = c(0, 0, 0, 1),
  j = c(1, 2, 3, 2),
  k = c(2, 3, 1, 3),
  facecolor = toRGB(viridisLite::viridis(4))
))
```

#### More complex tri-surface plot

We can use use `plotly` to create more complex tri-surf plot with proper data set. 
The following is an example, using the dataset to plot maps of given countries 
(North America for this specific example):

```{r}
data(wrld_simpl)

map1 <- subset(wrld_simpl,
               NAME %in% c("Canada", "Greenland", "Mexico", "United States"))
## DEL model (like TRI in silicate)
delmesh <-  anglr::globe(anglr::DEL(map1, max_area = 0.5))
mesh <- as.mesh3d(delmesh)


# plot point cloud
x <- mesh$vb[1,]
y <- mesh$vb[2, ]
z <- mesh$vb[3,]
m <- matrix(c(x,y,z), ncol=3, dimnames=list(NULL,c("x","y","z")))

# colours in z don't make sense here, need to map object aesthetics above
zmean <- apply(t(mesh$it),MARGIN=1,function(row){mean(m[row,3])})

library(scales)
facecolor = colour_ramp(
  brewer_pal(palette="RdBu")(9)
)(rescale(x=zmean))

suppressPlotlyMessage(plot_ly(
  x = x, y = y, z = z,
  i = mesh$it[1,]-1, j = mesh$it[2,]-1, k = mesh$it[3,]-1,
  facecolor = facecolor,
  type = "mesh3d"
))
```
## Conclusion

In this tutorial, we introduced how to use the **rgl** package and its'
relative functions to create 3D scatter plots, using **plotly** to create
interactive 3D surface plots and more complex tri-surface plots. We can see that
plotting scattered plots in 3D can be very useful such that we can fit a regression
plane for it for future analysis. Through **plotly** we can also create very smooth
3D interactive visualizations. However, it is not to say that 3D excels 2D plots,
in most cases we need to makes choices based on the specific case and scenario.
Sometimes 3D plots overcomplicates the pattern and information that could be
directly shown in 2D plots.


## References

[A complete guide to 3D visualization device system in
R](http://www.sthda.com/english/wiki/a-complete-guide-to-3d-visualization-device-system-in-r-r-software-and-data-visualization)

[rgl
Overview](https://cran.r-project.org/web/packages/rgl/vignettes/rgl.html#documents-with-rgl-scenes)

[Make beautiful 3D plots in R - An Enhancement to the
Storytelling](https://towardsdatascience.com/make-beautiful-3d-plots-in-r-an-enhancement-on-the-story-telling-613ddd11e98)

[Embed an interactive 3D plot with
rgl](https://bookdown.org/yihui/rmarkdown-cookbook/rgl-3d.html)

[Plotly Graphing Libraries](https://plotly.com/graphing-libraries/)
