# Visualizing Time Series Data

Kate Lassiter

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
suppressMessages(library(ggfortify))
library(gridExtra)
library(grid)
library(timevis)
suppressMessages(library(forecast))
suppressMessages(library(plotly))
library(tidyr)
library(zoo)
```

#### Starting Point
Same exploratory questions as with any new data set:\

* Strongly correlated columns
* Variable means 
* Sample variance, etc.\

Use familiar techniques:\

* Summary statistics
* Histograms
* Scatter plots, etc.

Be very careful of lookahead!\

* Incorporating information from the future into past smoothing, prediction, etc. when you shouldn't know it yet
* Can happen when time-shifting, smoothing, imputing data
* Can bias your model and make predictions worthless


#### Working with time series (ts) objects
Integration of ts() objects with ggplot2:\

* ggfortify package
  * autoplot()
  * All the same customizations as ggplot2
  * Don't have to convert from ts to dataframe format
* gridExtra package
  * Arrange the 4 ggplot plots as a 4-panel grid
* grid package
  * Add title to the grid arrangement

```{r,warnings=FALSE,messages=FALSE}
dax=autoplot(EuStockMarkets[,"DAX"])+
  ylab("Price")+
  xlab("Time")

cac=autoplot(EuStockMarkets[,"CAC"])+
  ylab("Price")+
  xlab("Time")

smi=autoplot(EuStockMarkets[,"SMI"])+
  ylab("Price")+
  xlab("Time")

ftse=autoplot(EuStockMarkets[,"FTSE"])+
  ylab("Price")+
  xlab("Time")
grid.arrange(dax,cac,smi,ftse,top=textGrob("Stock market prices"))

```

#### Time series relevant plotting:\
Working with the data:

* Directly transforming ts() objects for use with ggplot2:\
  * complete.cases() to easily remove NA rows - prevent ggplot warning
  * avoid irritations of working with ts objects

Looking at changes over time:

* Plot differenced values
  * Histogram/scatter plot of the lagged data
  * Shows change in values, how values change together 
  * Trend can hide true relationship, make two series appear highly predictive of one another when they move together
  * Use base package diff(), calculates difference between point at time _t_ and _t+1_
```{r,warnings=FALSE}
new=as.data.frame(EuStockMarkets)
new$SMI_diff=c(NA,diff(new$SMI))
new$DAX_diff=c(NA,diff(new$DAX))

p1 <- ggplot(new, aes(SMI,DAX))+
  geom_point(shape = 21, colour = "black", fill = "white")
p2 <- ggplot(new[complete.cases(new),], aes(SMI_diff,DAX_diff))+
  geom_point(shape = 21, colour = "black", fill = "white")

grid.arrange(p1,p2,top=textGrob("SMI vs DAX"))
```

Exploring Time Lags:

* Lagged differences:\
  * Time series analysis: focused on predicting future values from past
  * Concerned whether a change in one variable at time _t_ predicts change in another variable at time _t+1_
  * lag() to shift forward by one
  * Showing density using alpha
```{r}
new$SMI_lag_diff=c(NA,lag(diff(new$SMI),1))
ggplot(new[complete.cases(new),], aes(SMI_lag_diff,DAX_diff))+
  geom_point(shape = 21, colour = "black", fill = "white",alpha=0.4,size=2)
```

Now there is no apparent relationship: positive change in SMI today won't predict positive change in DAX tomorrow. There is a positive trend over the long term, but this does little to predict in the short term

Observations:\

* Careful with time series data: use same techniques, but reshape data
* Change in values from one time to another is vital concept 

### Dynamics of Time Series Data
#### Seasonality, Cycle, Trend

Three aspects of time series data:

* Seasonal: 
  * Recurs over a fixed period 
* Cycle:
  * Recurrent behaviors, variable time period
* Trend: 
  * Overall drift to higher/lower values over a long time period\

These can be gathered through visual inspection:\

* Line plot
  * Clear trend
    * Consider log transform or differencing
  * Increasing variance
  * Multiplicative seasonality
    * Seasonal swings grow along with overall values
```{r}
autoplot(AirPassengers)+
  xlab("Year")+
  ylab("Passengers")
```

* Time series decomposition:
  * Break data into seasonal, trend, and remainder components
  * Seasonal component: 
    * LOESS smoothing of all January values, February values, etc. 
    * Moving window estimate of smoothed value based on point's neighbors 
  * stats package
    * stl()\
    
```{r}
autoplot(stl(AirPassengers, s.window = 'periodic'), ts.colour = 'red')+
  xlab("Year")
```
  \
  
  * Observations
    * Clear rising trend
    * Obvious seasonality
    * Difference between the two methods:
      * This particular decomposition shows additive, not multiplicative seasonality
        * But start and end time series have highest residuals
        * Settled on the average seasonal variance
    * Both reveal information on patterns that need to be identified and potentially dealt with before                forecasting can occur
      
#### Plotting: exploiting the temporal axis
##### Gannt charts
  * Shows overlapping time periods, duration of event relative to others
  * timevis package
```{r}
dates=sample(seq(as.Date('1998-01-01'), as.Date('2000-01-01'), by="day"), 16)
dates=dates[order(dates)]
projects = paste0("Project ",seq(1,8)) 

data <- data.frame(content = projects, 
                    start = dates[1:8],
                    end = dates[9:16])
timevis(data)
```
##### Using month and year creatively in line plots
  * forecast package
    * ggseasonplot()
    * ggmonthplot()
    * suppressMessages() prevents printing information outputted when loading a package
    
```{r}
ggseasonplot(AirPassengers)
ggmonthplot(AirPassengers)
```

  * Observations
    * Some months increased more over time than others
    * Passenger numbers peak in July or August
    * Local peak in March most years
    * Overall increase across months over the years
    * Growth trend increasing (rate of increase increasing)\
\

##### 3-D Visualizations: plotly package\

  * Convert to a format plotly will understand
    * Avoid using ts() object
    * Dataframe with datetime, numeric columns
    * lubridate package for date manipulation
```{r}
new = data.frame(AirPassengers)
new$year=year(seq(as.Date("1949-01-01"),as.Date("1960-12-01"),by="month"))
new$month=lubridate::month(seq(as.Date("1949-01-01"),as.Date("1960-12-01"),by="month"),label=TRUE)
plot_ly(new, x = ~month, y = ~year, z = ~AirPassengers, 
             color = ~as.factor(month)) %>%
    add_markers() %>%
    layout(scene = list(xaxis = list(title = 'Month'),
                        yaxis = list(title = 'Year'),
                        zaxis = list(title = 'Passenger Count')))
```
  * Allows a better view of the relationships between month and year 


#### Data Smoothing
* Usually need to smooth the data before starting analysis or visualization
    * Allows better storytelling
    * Irrelevant spikes dominate the narrative
* Methods:/
  * Moving average/median
    * Good for noisy data
    * Rolling mean reduces variance
      * Keep in mind: affects accuracy, R² statistics, etc.
      * Zoo package rollmean()
      * Prevent lookahead, use past values as the window (align="right")
      * k = 6  is a 6 month rolling window
      * gsub() substitute series names for the clearer legend
      * tidyr package gather()
        * Convert from wide to long, use this as color/group in ggplot

```{r}
new = data.frame(AirPassengers)
new$AirPassengers=as.numeric(new$AirPassengers)
new$year=seq(as.Date("1949-01-01"),as.Date("1960-12-01"),by="month")

new = new %>%
  mutate(roll_mean = rollmean(new$AirPassengers,k=6,align="right",fill = NA))

df <- gather(new, key = year, value = Rate, 
                            c("roll_mean", "AirPassengers"))
df$year=gsub("AirPassengers","series",df$year)
df$year=gsub("roll_mean","rolling mean",df$year)
df$date = rep(new$year,2)

ggplot(df, aes(x=date, y = Rate, group = year, colour = year)) + 
  geom_line()
  
```
  * Exponentially weighted moving averages 
    * Weigh long past values less than recent
  * Geometric mean 
    * Combats strong serial correlation
    * Good for series with data that compounds greatly as time goes on
